---
title: "Categorical predictors"
author: "Marcin Włodarczak"
date: "`r format(Sys.time(), '%d %B %Y')`"
editor_options:
  chunk_output_type: console
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = FALSE, size = 'footnotesize', fig.align = 'center', out.width = '0.7\\textwidth', cache = FALSE)
```

```{r message=FALSE}
library('tidyverse')
theme_set(theme_bw())
```

# Categorical vs. numerical  predictors

So far we have dealt with numerical predictors but often we are interested in differences between *groups* of some kind. How to deal with those? Fortunately, the linear modelling strategy we have learnt can easily cover these cases.

# Two classes

Let's generate data for two groups (`A` and `B`), each with 30 data points, sampled from normal distributions with the means of 5 and 3 and the same standard deviation of 4:
```{r}
d <- tibble(score = c(rnorm(30, 5, 4), rnorm(30, 3, 4)),
            group = rep(c('A', 'B'), each = 30))
head(d)
tail(d)
```

Since we `group` only takes one of two values (`A` or `B`), it makes sense to convert it to a *factor*:
```{r}
d <- d %>%
  mutate(group = as_factor(group))
d$group
```

We can plot this data as boxplots:
```{r}
ggplot(d, aes(x = group, y = score)) +
  geom_boxplot()
```

The question now is how we can *model* this data. As it happens, we can simply plug the categorical variable into the regression formula:
```{r}
m1 <- lm(score ~ group, d)
summary(m1)
```

That's great but what do these coefficients mean? What R does under the hood is map one of the groups (by default the first one when group names are sorted alphabetically) to 0 (the intercept) and the other group to 1. Indeed we can see that:

- the intercept estimate (`r round(coef(m1)[1], 4)`) is equal to the mean of group `A`
- the slope (`r round(coef(m1)[2], 4)`) corresponds to the difference between the second group mean and the intercept

```{r}
d %>%
  group_by(group) %>%
  summarise(mean_score = mean(score))
```

Note that we can *re-level* the group factor and make `B` the reference using `fct_relevel`:
```{r}
d <- d %>%
  mutate(group = fct_relevel(group, 'B'))
levels(d$group)
```

If we refit the model now, the intercept will be equal to the mean of group `B` and the slope will the same but with the opposite sign:
```{r}
m2 <- lm(score ~ group, d)
coefficients(m2)
```

## Exercise

The data set `ratings` from the `langaugeR` package includes, among other things, average ratings of referent size for a bunch of English words, marked for whether the referent is an animal or a plant:
```{r}
library('languageR')
data('ratings')
```

Let's extract the `Word`, `Class` and `meanSizeRating` columns:
```{r}
ratings <- ratings %>%
  select('Word', 'Class', 'meanSizeRating')
head(ratings)
```

1. Visualize the distribution of average size ratings for the animal and plant word categories and calculate their mean values.
```{r}
# Write your solution here.
```

2. Fit a regression model predicting mean size rating from the category of the referent and interpret the model coefficients.
```{r}
# Write your solution here.
```

3. Refit the model with "Plant" as the reference.
```{r}
# Write your solution here.
```

# Three or more classes

The above can be easily generalized to situations when we have more than two classes:
```{r}
d3 <- tibble(score = c(rnorm(30, -5, 2), rnorm(30, 2, 2), rnorm(30, 8, 2)),
             group = rep(c('A', 'B', 'C'), each = 30))
```

Again, we will convert `group` to a factor:
```{r}
d3 <- d3 %>%
  mutate(group = as_factor(group))
```

Note that since `A` comes first in the alphabet, it will be used as the reference:
```{r}
levels(d3$group)
```

Here is the distribution of `score` in each group:
```{r}
ggplot(d3, aes(group, score)) +
  geom_boxplot()
```

Let's fit the model:
```{r}
m3 <- lm(score ~ group, d3)
summary(m3)
```

As before, the intercept is equal to the mean of the reference group (in our case, `A`). The other two coefficients are equal to the difference between the respective group means and the intercept:
```{r}
d3 %>%
  group_by(group) %>%
  summarise(mean_score = mean(score))
```

Note that we do not get the comparison between `B` and `C`. To do that we need to use one of them as a reference:
```{r}
d3 <- d3 %>%
  mutate(group = fct_relevel(group, 'B'))
m4 <- lm(score ~ group, d3)
summary(m4)
```

Now we have the `B`-`C` and `B`-`A` comparisons but not the `A`-`C` comparison.

## Exercise

Re-level the `group` factor to use `C` as a reference, refit the model and interpret the coefficients.
```{r}
# Write your solution here.
```

## Exercise

The file `lynott09.csv` contains a subset of data from Lynott, D. and Connell, L. (2009). *Behavior Research Methods* 41: 558. https://doi.org/10.3758/BRM.41.2.558, which measured the strength of association between adjectives and different modalities (auditory, haptic, olfactory etc.). **Remember to set the working directory to the location of the `Rmd` file before you attempt to read in the data.**
```{r message=FALSE}
lynott09 <- read_csv('lynott09.csv')
```

Create a model which predicts the degree to which adjectives with different dominant modalities (column `DominantModality`) are tied to a *single* perceptual modality (column `ModalityExclusivity`). Interpret the model coefficients.
```{r}
# Write your solution here.
```

# Revisiting the residuals

Remember analysis of residuals from last time. It is pretty much the same thing here.  First of all, we would like residuals to be distributed normally.
```{r}
res <- residuals(m4)
hist(res)
qqnorm(res)
qqline(res)
```

Second, we want to see them have the same variance for each group:
```{r}
fitted_vals <- fitted(m4)
plot(fitted_vals, res)
```

That looks alright but are the fitted values? Remember that the expected value for each group is simply its mean:
```{r}
d3 %>%
  group_by(group) %>%
  summarise(mean_score = mean(score))
```

Remember how we said at the very beginning that you can think of the mean as a very simple model? Here you can see that this is really the case!

## Exercise

Analyse the distribution of residuals in the `lynott09` model. 
```{r}
# Write your solution here.
```

# Combining categorical and continuous predictors

Obviously, we can combine numerical and categorical predictors in a single model. Below we revisit the `lexdec` lexical decision data (from the `languageR` package, see `?lexdec` for more information) again:
```{r}
data(lexdec)
```

Let's create a data summary with some statistics for each word:

- `Class`: category of the referent (`animal` or `plant`)
- `Frequency`: word frequency (log-transformed)
- `Length`: word length in characters
- `RT`: mean reaction time (in $\log(ms)$)

```{r}
words <- lexdec %>%
  group_by(Word) %>%
  summarise(Class = first(Class),
            Frequency = first(Frequency),
            Length = first(Length),
            RT = mean(RT))
```


Let's fit a regression model which predicts reaction time (`RT`) from word frequency (`Freq`) and word class (`Class`)
```{r}
lexdec_m <- lm(RT ~ Frequency + Class, words)
summary(lexdec_m)
```

Here is how we would interpret this model (remember that reaction time is expressed on a logarithmic scale!):

- Intercept: the expected reaction time when both predictors are equal to 0, that is for a word denoting an animal with a frequency of 0. We can tell that the `animal` class was mapped on the intercept because `Classplant` is present in the model summary and therefore is not used as a reference. In addition, "animal" precedes "plant" when sorted in alphabetical order.
- Frequency: the change in reaction time corresponding to an increase of one in word frequency (note that frequency is also expressed on a logarithmic scale).
- Class: The difference in reaction time between the `plant` and `animal` categories.

## Exercise 

Using the data above, standardize `Frequency` and refit the model. How do you interpret the coefficients now?

```{r}
# Write your solution here.
```

## Exercise

Re-fit the model above with word length (`Length`) as an additional predictor. Interpret the coefficients.

```{r}
# Write your solution here.
```


